\documentclass{ximera}
\title{Cycloidal Arclengths: An Animated Approach}

\newcommand{\pskip}{\vskip 0.1 in}

\newtheorem{theorem19}{Robinson's Arclength Theorem for Prolate Cycloids, Local Version}
\newtheorem{theorem20}{An Arclength Theorem for Prolate Epi and Hypocycloids}
\newtheorem{theorem21}{Robinson's Lemma}
\newtheorem{theorem22}{The Inclination Lemma}
\newtheorem{theorem23}{An Arclength Theorem for General Cycloids}

\begin{document}
\begin{abstract}
Relationship between their arclengths.
\end{abstract}
\maketitle
 
\section{Introduction}

%Figure 1 shows a \emph{prolate cycloid} (from the Latin \emph{poferre}, to extend and the Greek \emph{kyklos}, circle). It is the curve traced by a point $P_1$ outside a circle rolling on a line

Roll a circle ${\cal C}_1$ of radius $a$ once through its circumference on a line ${\cal L}$ and a point $P_1$ fixed in the reference frame of the circle sweeps out one arch of a \emph{general cylcoid}. When $P_1$ is outside the circle, the \emph{prolate cylcoid} (from the Latin \emph{poferre}, to extend and the Greek \emph{kyklos}, circle) crosses ${\cal L}$. In a recent article,  D. Robinson showed the one-arch length of the prolate cycloid on the same side of ${\cal L}$ as ${\cal C}_1$ exceeds the opposite-side length by $8a$. The excess surprisingly does not depend on the distance $b>a$ of the tracing point from the center of ${\cal C}_1$.  Choosing $b=a$ shows the excess equals the one-arch length of the \emph{ordinary} $a$-cycloid traced by a point $P_3$ on the circumference of the same rolling circle (see Figure 1, where ${\cal L}$ is the $x$-axis, $a=2$, and $b=3.6$).

\begin{onlineOnly}
    \begin{center}
\desmos{ty0hdz7uyj}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/ty0hdz7uyj}{Limacon Cylcoid Arclength 1B}

%What if $b<a$?. Then the tracing point lies inside the rolling circle and the \emph{curtate} (from the Latin \emph{curtus}, cut short) cycloid does not cross the $x$-axis. But it does have inflection points. And the convave-down arclength of one arch exceeds the concave-up length by $8b$ (Figure ??). Each of these arclength theorems follows quickly from the other. But it took us a long time to realize this and so we save this discussion for last (Section 7). %let alone to stumble upon the second theorem. So we save this discussion for last (Section 7).

If instead the tracing point lies $a<b$ units from the center of the rolling circle (radius $b$),  the \emph{curtate} (from the Latin \emph{curtus}, cut short) cycloid does not cross the $x$-axis. But it does have inflection points. And the convave-down arclength of one arch still exceeds the concave-up length by $8a$ (Figure ??). Each of these arclength theorems follows quickly from the other. But it took me a long time to realize this and so I save this discussion for last (Section 7).

\begin{onlineOnly}
    \begin{center}
\desmos{0eosvsfm9r}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/0eosvsfm9r}{Curtate Cycloid 33}

 
Robinson proves two other similarly surprising facts. One is about the limacon with polar equation $r=A + B\cos\theta$, $\theta\in [0,2\pi)$. When $B>A$ the limacon intersects itself and the length of its outer loop exceeds the length of the inner loop by $8A$. Taking $B=A$ shows the excess equals the length of the cardioid $r=A + A\cos\theta$.

The other is about ellipses. It says when $B\geq A$ and 
\[
      \alpha = \frac{1}{2} \cos^{-1} \left( \frac{A-B}{A+B}  \right)  =  \sin^{-1}\left( \frac{B}{A+B}  \right) ,
\]
the arclength of the quater-ellipse
\[
   (x,y) = (A \cos\theta , B\sin\theta) \, , \, 0\leq \theta \leq \pi/2 ,
\]
over the interval $\theta \in [0,\alpha]$ exceeds its arclength over the interval $\theta\in [\alpha, \pi/2]$ by $B-A$.

Robinson's proofs exploit the commonality these three types of curves have in their arclength functions being elliptic integrals. But the curves have another commonality. They are all \emph{roulettes} (from the Latin \emph{rota} for wheel) traced by a point fixed in the reference frame of a rolling circle.

%A point $b$ units from the center of a circle with radius $a$ sweeps out the limacon $r=2a + 2b\cos\theta$ as it rolls once around a circle with the same radius. 

The limacon $r=A+B\cos\theta = 2a + 2b \cos\theta$ is traced by a point $b$ units from the center of a circle with radius $a$ rolling around the outside of a circle with the same radius. And (surprisingly, at least for me) a point $b>a$ units from a circle with radius $a$ rolling around the \emph{inside} of a circle with radius $2a$ sweeps out an ellipse with semi-major and minor axes $b+a$ and $b-a$.

These observations suggest Robinson's arclength equalities are special cases of a more general theorem about epicycloids and hypocycloids (curves traced by a point fixed in the reference frame of one circle as it rolls around the outside or inside of another). For the limacon $r= 2a+2b$, the excess arclength $8(2a) = 16a$ is the length of the cardioid traced by a point attached to a circle with radius $a$ as it rolls once around a congruent circle. It is not a coincidence that the length of the cardioid
is twice the length of the cycloid traced by a point attached to the same circle as it rolls on a line. Replacing the line with a circle of radius $a$ \emph{doubles} corresponding arclengths of the roulette.

And for the semi-ellipse 
\[
    (x,y) = (A\cos\theta, B\sin\theta) =  ((b-a)\cos\theta , (b+a)\sin\theta) \, , \, 0\leq \theta \leq \pi, 
\]
traced by a point $b$ units from the center of a circle with radius $a$ as it rolls \emph{once} through its circumference around the inside of a circle of radius $2a$, the excess is
\[
     2(B-A) = 4a . 
\]
When $b=a$ (ie. when $A=0$), the tracing point (now on the rolling circle with radius $a$) traces a diameter of the fixed circle with radius $2a$. It is also not a coincidence that the length of this roulette is half the length of the cycloid. Replacing the line with a circle of radius $2a$ halves arclengths when the smaller circle rolls around the inside of the larger. We explore these ideas in Sections 5 and 6. 

%And what about \emph{curtate} (from the Latin \emph{curtus}, cut short) cycloids, curves traced by points inside a rolling circle? It seems like there should be an arclength theorem for these, and in fact there is (Section??) 

Near the end of his paper Robinson asks for a heuristic explanation of his arclength theorem and the genesis of this article was to find one. While we fall short of this mark, a kind proof-without-words in a single picture, we do offer an animated proof with minimal computation (Section 4). 


%offers some geometric insight into why these arclength theorems are true. %By taking a geometric approach (Sections ????) and working locally, we keep computations to a minimum. %  by avoiding integrals and parameteric descricptions of curves  %, keep computations to a minumum and avoid relying both on integration and parametric descriptions of the curves.


% and the genesis of this article was to find such a proof. While we fell short of the mark, a kind of proof-without-words in a single picture, we do take a more geometric approach to prove a more general theorem about arclengths of prolate cycloids (Sections ??-??), a proof that avoids elliptic integrals  and keeps the computations to a miniumum. We hope it offers some insight into why these theorems are true.



%The purpose of this note is to give a geometric proof of this result and some of Robinson's other discoveries.  %this fact and view some of Robinson's other results in a broader context.

\section{An Arclength Theorem for Prolate Cycloids}

Our first step in proving Robinson's arclength thoeorem was to look for a more general equality of arclengths by associating each point of a prolate cycloid above the $x$-axis with a point below. The simplest choice was to pair points with same inclination angle (the angle a curve makes with the horizontal) and experiments with desmos suggested the same equality of arclengths between any two inclination angles.

In the figure below, for example, the difference $s_1 - s_2 = A_1P_1 - A_2P_2$ in the same and opposite-side arclengths of the $(a,b)$ prolate cycloid between inclination angles $0$ and $\phi$ equals the arclength $s_3  = A_3P_3$ of the ordinary $a$-cycloid between these same inclination angles. 

\begin{onlineOnly}
    \begin{center}
\desmos{0vodc6gvp5}{900}{600}   %bpvuhnhi35
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/0vodc6gvp5}{Limacon Cylcoid Arclength 2B}

You can almost see the equality $s_1 - s_2 = s_3$ of small arclengths in the figure below by looking at parallel arcs of the same color between the same inclination angles.

%The figure below makes the equality of arclengths plausible if you focus on the short arcs of the same color between the same inclination angles.

\begin{onlineOnly}
    \begin{center}
\desmos{yqw8ektvps}{900}{600}     %4xn0xtuszi
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/yqw8ektvps}{Cylcoid Trochoid Color 2}

%These equalities of arclengths gives a local version of Robinson's theorem. This local version below pairs

\begin{theorem19}
%As in Figure 2, let $s_1 = s_1(\phi)$ be the arclength of a prolate cycloid on the same side of the $x$-axis as its rolling circle between inclination angles $0$ and $\phi$, $0\leq \phi \leq \pi/2$. Similarly, let $s_2 = s_2(\phi)$ be the opposite-side arclength of the prolate cycloid and $s_3 = s_3(\phi)$ the arclength of the sister cylcoid, both between inclination angles $0$ and $\phi$.

Let $s_1 = s_1(\phi)$ be the same-side arclength of the $(a,b)$ prolate cycloid between inclination angles $0$ and $\phi$, $0\leq \phi \leq \pi/2$. Let $s_2 = s_2(\phi)$ be the opposite-side arclength of the same cycloid between inclination angles $\pi - \phi$ and $\pi$ . And let $s_3 = s_3(\phi)$ be the arclength of the ordinary $a$-cylcoid between inclination angles $0$ and $\phi$. Then
\[
s_1 - s_2 = s_3.
\]
\end{theorem19}

Trying to prove this more general result worked to our advantage as its shifted our focus from a global one (proving an equality of integrals) to  a local one (proving an equality of \emph{differential} arclengths). Our proof is primarily geometric and relies on the instantaneous center of rotation.

\emph{Remark:} The theorem measures the inclination angle to be a strictly increasing function of arclength. While in Figure 2, the arclength $s_2=A_2P_2$ is between inclination angles $\pi$ and $\pi +\phi$ and lies outside one arch of the problate cycloid (where $-\pi \leq \phi \leq \pi$, as in Figure 1), it does equal the arclength between inclination angles $\pi-\phi$ and $\pi$. Replacing the interval $[\pi-\phi, \pi]$ with $[\pi, \pi +\phi]$ makes it easier to illustrate the proof of Robinson's theorem (see Figure ?? of Section 3).





\section{The Instantaneous Center of Rotation}
From the \emph{CarTalk} puzzler archives, May 15, 2006:

\small{{\bf RAY:} A car is traveling at 60 mph. You're standing by the side of the road, or lying on the road as the case may be. Can you name a part, or parts, of the car, that are not moving in relation to the road?

If needed, x-ray vision is permitted.

That's Part A.

Part 2-- and this is a hint-- what part of the car is traveling at 120 mph?}

%{\bf Answer:}

%The bottom of every tire is moving at zero miles an hour with respect to the road. If that weren't the case, then the car would be in a skid.

%Imagine an ideal tire where one point is touching the road. That very point of the tire that's touching the road must not be moving in relation to the road. In fact, it is moving at zero miles an hour. And when you point it 180 degrees away, it has to be moving at twice that speed, or 120 miles an hour.

\href{https://www.cartalk.com/radio/puzzler/special-non-moving-car-part}{The Special, Non-Moving Car Part}



It might seem counterintuitive that at any moment a rolling wheel is rotating \emph{not} about its center but about its point of contact with the ground. This point, the wheel's \emph{instantaneous center of rotation}, is the point in the reference frame of the rolling wheel momentarily at rest (In Figure 1 the point $P_3$ is the wheel's center of rotation at the cusps, where it comes to a stop and reverses direction.) And at any moment, each point $P$ at rest in the reference frame of the rolling wheel,  like a pebble stuck in a tire, is rotating about this center. This has two implications.         %The cusps of the ordinary cycloid in Figure 1 reveal why, because as the tracing point bounces off the $x$-axis it reverses direction and comes to a stop.

%At any moment, each point $P$ at rest in the reference frame of the rolling wheel,  like a pebble stuck in a tire, rotates about this center of rotation.  This has two implications. 

\begin{enumerate}
\item The speed of $P$ is equal to the product $\omega \rho$ of the wheel's rotation rate and the \emph{rotation arm} $\rho = PQ$ (the distance from $P$ to the center of rotation).

\item The velocity of $P$ is orthogonal to $\overline{PQ}$.
\end{enumerate}

%The photograph of a rolling wheel below gives some sense of (a); points on the spokes farther from the point of contact are moving faster and the spokes blur....

The figure below illustrates the motion of a point $P$ attaced to the circumference of a wheel rolling along a straight road at a consant speed. The marked angle $\theta$ is the wheel's rotation angle measured from an arbitrary direction, in this case from the upward-vertical when $P$ coincides with the top point $A$ of the cycloid. Because of (b) above, the incination of the cycloid at $P$ (ie. the angle between the tangent and the horizontal) is equal to the measure of $\angle BQP$ (ie. the angle between the normal $\overline{PQ}$ and the vertical). The points on the cycloid are spaced at equal time intervals and suggest the speed $v= \rho (d\theta/dt)$ of the tracing point is proportional to the rotation arm $PQ$. 

 %shows the ordinary cycloid traced by a point $P$ stuck on the circumference of a rolling wheel. The marked angle $\theta$ is the wheel's rotation angle measured from an arbitrary direction, in this case from the upward-vertical when $P$ coincides with the top point $A$ of the cycloid. Because of (b) above, the incination of the cycloid at $P$ (ie. the angle between the tangent and the horizontal) is equal to the measure of $\angle BQP$ (ie. the angle between the normal $\overline{PQ}$ and the vertical). The points on the cycloid are spaced at equal intervals of rotation and suggest the speed $v= \rho (d\theta/dt)$ of the tracing point is proportional to the rotation arm $PQ$. %its distance from the center of rotation.  

\begin{onlineOnly}
    \begin{center}
\desmos{ilinb0qn7q}{900}{600}   %bpvuhnhi35
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/ilinb0qn7q}{Center of Rotation}

%To prove the arclength theorem, it helps to think differentially. Then (a) is equivalent to saying

Locally (a) says that as the wheel turns through the differential angle $d\theta>0$, the tracing point traverses the differential distance $ds = \rho \, d\theta$. For the $a$-cycloid above, for example, traced by a point attached to a circle of radius $a$,
\begin{align*}
     ds = &\rho \, d\theta \\
             &= 2a \cos\phi\, d\theta \\
              &= 2a \cos \phi \, (2 \, d\phi) .  
\end{align*}

{\bf Remarks:} 
\begin{enumerate}
\item Adding these differential arclengths shows the length of the cycloid from $A$ to $P$ 
\[
      s =\int_0^\phi 4a\cos\phi \, d\phi = 4a\sin\phi                   % s = \int_0^\theta (2 4a\sin(\theta/2) = 4a \sin\phi %2 \text BP .
\]
is twice the length of chord $\overline{BP}$.

My favorite way to prove this equality is to compare the differential arclength $ds = 2a\cos\phi\, d\theta$ with the differential change $dc$ in the distance $c = BP$. In the reference frame of the rolling circle where $P$ is at rest, $B$ traverses the differential arclength $a\, d\theta$ as the circle turns through the angle $d\theta$. But since $B$ moves in a direction inclined at the angle $\phi$ to $\overrightarrow{PB}$ %(and not directly away from $P$), 
\[
    dc = (a\, d\theta) \cos \phi = ds/2.
\]

\item Ray's answer to the puzzler:

\small{The bottom of every tire is moving at zero miles an hour with respect to the road. If that weren't the case, then the car would be in a skid.

Imagine an ideal tire where one point is touching the road. That very point of the tire that's touching the road must not be moving in relation to the road. In fact, it is moving at zero miles an hour. And when you point it 180 degrees away, it has to be moving at twice that speed, or 120 miles an hour.}

The last remark needs justification, but follows from the observation that the arithmetic mean (or more generally, the weighted average) of the velocities of two points at rest in the reference frame of the rolling wheel is equal to the velocity of the mean (or the same weighted average) of these points.

Rolling a book along the top of a cylinder is a convincing way to demonstrate that a pebble stuck in a tire moves twice as fast as the axle when it reaches the top of its path (and is twice as far from the center of rotation as the axle).  See Figure ??. %As illustrated in Figure ?, you can see the book advance twice as far as the cylinders.

%Roll a book on the tops of two cyinders (oatmeal containers work well) and you can see the book advance twice as far as the cylinders, a convincing way to demonstrate a pebble stuck in a tire moves twice as fast as the axle when it reaches the top of the wheel (and therefore twice as far from the center of rotation as the axle). %that at the top of its motion a pebble stuck in a tire moves twice as fast as the axle. 


\begin{onlineOnly}
    \begin{center}
\desmos{7miak40vna}{900}{600}   %bpvuhnhi35
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/7miak40vna}{Rolling Quaker Oats}

%A pebble stuck in the wheel of a tire moves twice as fast as the wheel's axle at the top of its motion. For a dramatic demonstration, roll a book along two cylinders (oatmeal containers work well) and you can see the book advance twice as far the cyclinders.

\end{enumerate}


\section{A Proof of Robinson's Theorem}

\begin{proof}
Animate Figure ?? and you'll see circles ${\cal C}_i$, $i=1,2,3$ roll along the $x$-axis as the points $P_i$ sweep out the same and opposite-side arclengths of an $(a,b)$ prolate cycloid and an ordinary $a$-cycloid.  Look carefully and you'll notice the rotation arms $\rho_i = Q_iP_i$ make the same angle $\phi$ with the vertical at all times. But while the circle for the ordinary cycloid rotates at a constant rate, the others do not. The key to our proof of Robinson's arclength theorem is to relate these rotation rates. 


%To prove the arclength theorem, imagine three circles ${\cal C}_i$, $i=1,2,3$, rolling along the $x$-axis as the points $P_i$ (at rest in the reference frames of these circles) respectively sweep out the upper and lower parts of the prolate cycloid and its sister cycloid.  Imagine also the rotation arms $\overline{Q_iP_i}$ make the same angle $\phi$ with the vertical at all times as in Figure 3. In the animated version of this figure, the circle for the ordinary cycloid (centered at $C_3$) rotates at a constant rate, but the others do not. The key to our proof, and to creating the animation, is to relate the rotation rates of the three circles. 


%We suppose the circles start with the points $P_i$ at the same inclination angle $\phi=0$ and the segments $Q_iP_i$ vertical. And we also assume as illustrated in Figure 3, the segments $\overline{Q_iP_i}$ from the centers of rotation to the tracing points remain parallel at all times.


\begin{onlineOnly}
    \begin{center}
\desmos{qtse3eivit}{900}{600}   %snq3rgm0a4
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/qtse3eivit}{Limacon Cylcoid Arclength 3C}

To prove the arclength theorem, we work locally and show
\begin{equation} \label{Eq:DiffArcs}
      ds_1 - ds_2 = ds_3 ,
\end{equation}
where  $ds_i$ are the differential arclengths along the respective curves between inclination angles $\phi$ and $\phi + d\phi$. 
A key point in what follows (see Figure ?? above) is that %by the symmetry of the first (red) circle about the diameter perpendicular to $\overline{P_1R_1}$,
\[
   \rho_2 = Q_2P_2 = Q_1R_1 = S_1P_1 ,
\]
where the last equality follows from the symmetry of ${\cal C}_1$ about the diameter perpendicular to chord $\overline{P_1R_1}$. This equality has three consequences:

\begin{enumerate}
\item  $\rho_1 - \rho_2 = \rho_3$ ,

\item $\theta_1 + \theta_2 = \theta_3$, where the rotation angles $\theta_1$ and $\theta_3$ are measured from the upward vertical and $\theta_2$ from the downward vertical as illustrated above, and %the rotation angles are measured from the vertical,  as indicated above) , and

\item $\frac{d\theta_1}{d\theta_2} = \frac{\rho_1}{\rho_2}$.
\end{enumerate}

The first (a) is immediate, but it alone does not imply the equality of differential arclengths (\ref{Eq:DiffArcs}). This is because the circles turn through different differential angles $d\theta_i$ in sweeping out these arclengths. We need also (b) and (c) to prove (\ref{Eq:DiffArcs}), which we now write in the form 
\[
       \rho_1 \, d\theta_1 - \rho_2 \, d\theta_2 = \rho_3 \, d\theta_3 .
\]


Assuming (a)-(c), 
\begin{align*}
     ds_1 - ds_2 &= \rho_1 \, d\theta_1 - \rho_2 \, d\theta_2 \\
                       &= (\rho_3 + \rho_2) \, d\theta_1 - (\rho_1 - \rho_3)\, d\theta_2 \\
                       &= \rho_3 (d\theta_1 + d\theta_2) + (\rho_2 \, d\theta_1 - \rho_1\, d\theta_2) \\
                       &= \rho_3 \, d\theta_3 \\
                       &= ds_3 .
\end{align*}

We are left with proving (b) and (c).     %It remains to prove (b) and (c).

The proof of (b) follows from the figure below that records the three rotation angles $\theta_i$ and the inclination angle $\phi$ in one static (non-rolling) circle. As the circles in Figure ?? (the previous figure) roll in tandem, the tracing points $P_i$ move around the static circle as chord $\overline{P_1P_3P_2}$ turns at a constant rate.

Now since $\overline{P_3P_1}\cong \overline{Q_1P_2}$, we know $\Delta C_1P_3P_1\cong \Delta C_1Q_1P_2$. Therefore, $m\angle P_1C_1P_3 = \theta_2$ and 
\[
 \theta_3 = m \angle B_1C_1P_3 = \theta_1 + \theta_2.
\]


\begin{onlineOnly}
    \begin{center}
\desmos{txbihum5r3}{900}{600}   
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/txbihum5r3}{Cylcoids Proof of Theorem 1}

To prove (c), we borrow an idea from Tabachnikov's \emph{Geometry and Billiards} and consider the differential changes $b\, d\theta_i = P_iP_i^\prime$, $i = 1,2$  in the respective arclengths $b\theta_i$ as chord $\overline{P_1P_2}$ turns through the differential angle $d\phi$.  

\begin{onlineOnly}
    \begin{center}
\desmos{ffj1pwg96n}{900}{600}         %xvwr9mhuvw
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/ffj1pwg96n}{Cylcoids Proof of Theorem 1CC} %i2050p22ud

%i2050p22ud

In Figure ?? above, the differential arcs $ZP_1^\prime$ and $WP_2^\prime$ (circular arcs centered at $Q_1$) have lengths
\[
    ZP_1^\prime = \rho_1 \, d\phi
\]
and
\[
  WP_2^\prime =  \rho_2 \, d\phi.
\]

But since a chord makes congruent angles with a circle at its endpoints, the differential right triangles $\Delta P_1ZP_1^\prime$ and $\Delta P_2 W P_2^\prime$ are similar. Hence,
\[
      \frac{P_1P_1^\prime}{P_2 P_2^\prime}  =  \frac{ZP_1^\prime}{WP_2^\prime} 
\]
and
\[
    \frac{b\, d\theta_1}{b \, d\theta_2} = \frac{ \rho_1 \, d\phi}{ \rho_2 \, d\phi} .
\]
%and (b) follows.
 
\end{proof}

{\bf Remark:} Writing (\ref{Eq:DiffArcs}) as
\[
   \frac{ds_1}{d\phi} - \frac{ds_2}{d\phi} = \frac{ds_3}{d\phi}
\]
gives another interpretation of the arclength theorem. It expresses the radius of curvature 
\[
  \frac{ds_3}{d\phi} = 2 \rho_3
\]
of the ordinary cycloid as a difference 
\[
   \frac{ds_1}{d\phi} - \frac{ds_2}{d\phi} = \frac{2\rho_1^2}{\rho_1+\rho_2} - \frac{2\rho_2^2}{\rho_1+\rho_2}
\]
in the radii of curvature of the same and opposite-side arcs of the prolate cycloid, all at the same inclination angle.%terms of as a relationship
%\[
%     r_1 - r_2 = r_3
%\]
%among the radii of curvature of the upper/llower prolate and ordinary cycloids at the same inclination angle.





\section{Prolate Epi and Hypocycloids}
A classic problem asks how many rotations one quarter makes as it rolls once around another. 

The rolling quarter picks up an extra rotation and makes two rotations as it rolls once around the other. This is because as the quarter rolls through its circumference it makes one clockwise rotation in the reference frame of the rolling tangent line as this tangent makes one clockwise rotation around the fixed quarter.

%As the quarter rolls through its circumference, it makes one clockwise rotation in the reference frame of the rolling tangent line (Figure ??). But since this tangent makes one clockwise rotation around the fixed quarter, the quarter makes not one but two rotations as it rolls once around the other. 


\begin{onlineOnly}
    \begin{center}
\desmos{0dlaaalmjl}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/0dlaaalmjl}{Rolling Quarter 2}

%The rolling quarter picks up an extra rotation and makes two rotations as it rolls once around the other. This is because as the quarter rolls through its circumference it makes one clockwise rotation in the reference frame of the rolling tangent line


The same is true locally. As the rolling quarter turns through the differential angle $d\theta$ in the reference frame of the rolling tangent, it rotates through the differential angle $d\theta^* = 2\, d\theta$ in the reference frame of the fixed quarter. A consequence is that the arclength of a cardioid is twice that of its sister cycloid, both globally and locally. These curves are traced by points attached to a circle as it rolls once through its circumference; along a line for the cycloid and around a circle of the same radius for the cardioid.  

To see why, attach a tracing point $P^*$ to the top rolling quarter of Figure 3 (see Figure 4). This point sweeps out a cardioid in the reference frame of the fixed quarter and an ordinary cycloid in the reference frame of the rolling tangent. Now as the quarter rolls through the differential arclength $d\sigma$ of its circumference it turns through the angles $d\theta = d\sigma/a$ in the reference frame of the rolling tangent and $d\theta^* = 2\, d\theta$ in reference frame of the fixed quarter. Hence, as the quarter rolls through the differential arclength $d\sigma$, $P^*$ sweeps out a differential arclength
%\[
 %  ds = Q^*P^* \, d\theta
%\]
\[
  ds^* = Q^*P^*\, d\theta^* =  2Q^*P^* \, d\theta  %= 2\, ds
\]
along the cardioid equal to twice the differential arclength $ds = QP\, d\theta$ along the cycloid. And if, as illustrated in Figure 4, we measure the inclination angle $\phi = angle P^*Q^*C^*$ of the cardioid \emph{relative to the rolling tangent line}, then $ds^* = 2\, ds$ between the common inclination angles $\phi$ and $\phi + d\phi$ of the two curves.


\begin{onlineOnly}
    \begin{center}
\desmos{grcnlkuqan}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/grcnlkuqan}{Rolling Quarter and Cardioid 2}


A similar relationship holds more generally. As a circle ${\cal C}$ of radius $a$ rolls through the differential arclength $d\sigma$ of its circumference around the outside of a fixed circle ${\cal C}_r$ with radius $r$, it turns through the angle $d\theta = d\sigma/a$ relative to the rolling tangent line. At the same time, the tangent turns with the same sense through the angle $d\sigma/r$ about the center of ${\cal C}_r$. So ${\cal C}$ turns through the angle
\begin{align}
     d\theta^* &= \left(\frac{1}{a} + \frac{1}{r} \right) \, d\sigma  \notag \\
                    &= \left( 1 + \frac{a}{r}  \right) \, d\theta .  \label{Eq:RollAngle}
\end{align}
in the reference frame of ${\cal C}_r$. The same equation also holds if we take $r$ to be negative when ${\cal C}$ rolls on the inside of ${\cal C}_r$.

In either case, the equation implies a similar relationship 
\[
  ds^*  =  \left( 1 + \frac{a}{r}  \right)  ds
\]
between the differential arclength $ds$ of a general $(a,b)$-cycloid (traced by a point $b$ units from the center of a circle of radius $a$ rolling on a line) and the differential arclength $ds^*$ of the $(a,b,r)$-epicycloid (or hypocycloid if $r<0$) traced by the same point on the same circle rolling around the outside (inside) of a circle of radius $r$. Here we measure the arclengths between inclinations angles $\phi$ and $\phi+d\phi$, where as for the cardioid, the inclination angle of the epi (hypo) cycloid is relative to the rolling tangent.


An immediate consequence is an arclength theorem for prolate epicycloids (hypocycloids), provided we define the \emph{same} (respectively \emph{opposite}) side arclengths to lie on the same (resp. opposite) side of the rolling tangent line as the rolling circle.

\begin{theorem20}

Let $s_1^* = s_1^*(\phi)$ be the same-side arclength of an $(a,b,r)$ prolate ($b>a$) epi or hypocycloid between inclination angles $0$ and $\phi$, $0\leq \phi \leq \pi/2$. Let $s_2^* = s_2^*(\phi)$ be the opposite-side arclength of the same curve between inclination angles $\pi-\phi$ and $\pi$. And let $s_3^* = s_3^*(\phi)$ be the arclength of the ordinary $(a,a,r)$ epicylcoid (or hypocycloid) between inclination angles $0$ and $\phi$. Then
\[
       s_1^* - s_2^* = s_3^* .
\]

%Let $s_1^*=s_1^*(\phi)$ and $s_2^* = s_2^*(\phi)$ be respectively the same and opposite side arclengths of an $(a,b,r)$ prolate ($b>a$) epi or hypocycloid between inclination angles $\phi$ and $\phi + d\phi$. %, the former on the same side of the tangent line as the rolling circle, the latter on the opposite side. 

%Let $s_3 = s_3(\phi)$ be the arclength of the sister $(a,a,r)$ epicylcoid (or hypocycloid) between the same inclination angles. % traced by a point attached to the circumference of a same rolling circle in the reference frame of the rolling tangent.

\end{theorem20}

\begin{proof}
By the arclength theorem for cycloids,
\[
    ds_1 - ds_2 = ds_3
\]
for the same and opposite-side differential arclengths $ds_1$, $ds_2$ of the $(a,b)$ prolate cycloid and $ds_3$ of its sister cycloid between inclination angles $\phi$ and $\phi + d\phi$. Multiplying both sides of the above equation by the factor $a/r + 1$ shows
\[
    ds_1^* - ds_2^* = ds_3^*
\]
and proves the theorem.
\end{proof}



Figure 5 illustrates an example with the $(a,b,r) = (2,3.6,4)$ prolate epicycloid. The ordinary $(2,2,4)$ epicycloid is a nephroid, traced by a point on a circle rolling around the outside of a circle with twice the radius. The same-side (red) arclength of the prolate epicycloid exceeds its opposite-side (blue) length by the length of the nephroid (dashed). The figure shows the rolling circle at a transition point between the same and opposite-side arcs when $P^*$ crosses the rolling tangent.

\begin{onlineOnly}
    \begin{center}
\desmos{czhdwjarf3}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/czhdwjarf3}{Wrapping Cycloids around Circles Nephroid}



Figure ?? shows an example with a single transition point, crossed twice.  Here a circle rolls once around another of the same radius. The ordinary $(a,a,a)$ epicycloid is a cardioid and the $(a,b,a)$ prolate epicycloid a limacon with polar equation $r=2a+2b\cos\theta$ (the pole is at the node and the polar axis points upward). Because the transition point is the node we recover Robinson's result. The limacon's outer arclength exceeds the length of its loop by the length $2(8a) = 16a$ of the ordinary $a$-cardioid equal to twice that of the ordinary $a$-cycloid.

% The lower arc is the loop of the limacon. So the epicyloid theorem says is the lower arc and so the limacon's outer arclength exceeds the loop's by the length $2(8a) = 16a$ of the limacon's sister cardioid equal to twice that of the ordinary $a$-cycloid). 
\begin{onlineOnly}
    \begin{center}
\desmos{kuhagddz3r}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/kuhagddz3r}{Wrapping Cycloids around Circles Limacon}


%Figure ?? shows an example closely related to the one in Figure ???. Here a circle ${\cal C_1}$ with radius $a$ rolls around the \emph{inside} of a circle ${\cal C}_1^\prime$ with radius $2a$. The point $P^*$, $b$ units from the center of ${\cal C}_1$, traces an ellipse with major and minor semi-axes $b+a$ and $b-a$. And as ${\cal C}_1$ rolls once round ${\cal C}_1^\prime$, the point $P_1$ (on ${\cal C}_1$) traces the horizontal diameter of ${\cal C}_1^\prime$ twice. Our theorem for prolate hypocycloids then implies the same-side arclength (red) of the ellipse exceeds its opposite-side length (blue) by $4a$, twice the diameter of ${\cal C}_1^\prime$. The figure shows $P^*$ at a transition point where it crosses the rolling tangent.


%\begin{onlineOnly}
 %   \begin{center}
%\desmos{yillqxp8mg}{900}{600}   %yfoczp8y2g
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/yillqxp8mg}{Wrapping Cycloids Around Circles Ellipse 2B}

%To better understand this result geometrically, we need to describe the transition points in another way and Robinson does just that. He shows these points divide the segments tangent to the ellipse, those with their endpoints on the extended axes, into segments $P^*B = b+a$ and $P^*A = b-a$, with lengths respectively equal to the semi-major $OX$ and semi-minor axes $OY$ (see Figure ???). To see why, we turn to a more practical way of drawing an ellipse.


%\begin{onlineOnly}
%    \begin{center}
%\desmos{2afbtinzyu}{900}{600}   %yfoczp8y2g
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/2afbtinzyu}{Wrapping Cycloids Around Circles Ellipse 2BB}


\section{Ellipses and the Trammel of Archimedes}

Roll a circle ${\cal C}_1$ with radius $a$ around the \emph{inside} of a circle ${\cal C}_1^\prime$ with radius $2a$. Then the tracing point $P^*$, still $b$ units from the center of ${\cal C}_1$, traces an ellipse with major and minor semi-axes $b+a$ and $b-a$ (Figure ??). And as ${\cal C}_1$ rolls once round ${\cal C}_1^\prime$, the point $P_1$ (on ${\cal C}_1$) makes one round trip along the horizontal diameter of ${\cal C}_1^\prime$. So our theorem for prolate hypocycloids implies the same-side arclength (red) of the ellipse exceeds its opposite-side length (blue) by $4a$, twice the diameter of ${\cal C}_1^\prime$. %Figure ?? shows $P^*$ at a transition point where it crosses the rolling tangent. 

%\begin{center}  
%\youtube{watch?v=SZ6_jjfqNBo}  
%\end{center}

%\href{https://www.youtube.com/watch?v=SZ6_jjfqNBo}{Tom Silva drawing an ellipse}

\begin{onlineOnly}
    \begin{center}
\desmos{yillqxp8mg}{900}{600}   %yfoczp8y2g
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/yillqxp8mg}{Wrapping Cycloids Around Circles Ellipse 2B}



%Figure ?? shows $P^*$ at a transition point where it crosses the rolling tangent. But Robinson gives a more insightful description. He shows the transition points divide the segments tangent to the ellipse, those with their endpoints on the extended axes, into segments $P^*B = b+a$ and $P^*A = b-a$, with lengths respectively equal to the semi-major $OX$ and semi-minor axes $OY$ (see Figure ???). To see why, we turn to a more practical way of drawing an ellipse, one discovered some two millenia ago but still in use today.

Figure ?? shows $P^*$ at a transition point where it crosses the rolling tangent and the rotation arm $\overline{Q_1P^*}$ is perpendicular to the radius $\overline{C_1Q_1}$. But Robinson gives a better description of these points.

\begin{theorem21}
The transition points divide the segments tangent to the ellipse, those with their endpoints on the extended axes, into segments $P^*B = b+a$ and $P^*A = b-a$, with lengths respectively equal to the ellipse's semi-major $OX$ and semi-minor axes $OY$ (see Figure ???).
\end{theorem21}

\begin{onlineOnly}
    \begin{center}
\desmos{2afbtinzyu}{900}{600}   %yfoczp8y2g
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/2afbtinzyu}{Wrapping Cycloids Around Circles Ellipse 2BB}

To understand why, we turn to a more practical way of drawing an ellipse, one discovered some two millenia ago but still in use today.

In an episode of \emph{Ask This Old House} Tom Silva uses the trammel of Archimedes to construct an ellipse. A \emph{trammel} is a restraint or hindrance, and restraining two points on a thin rod (the trammel's arm) to slide along two usually perpendicular tracks forces any third point on that rod to trace an ellipse. A trammel of Archimedes hides in Figure ??. As ${\cal C}_1$ rolls around the inside of ${\cal C}_1^\prime$, the endpoints of the diameter $\overline{P_1R_1}$ through the tracing point slide along the axes of the ellipse (see Figure ?? for an explanation. CAPTION: Because an angle inscribed in a semi-circle a right angle, the normals to the moment arms $\overline{QP_i}$ pass through $O$. So points on the rolling circle always move directly toward or away from $O$. Each therefore moves along a diameter.) This trammel, with arm $\overline{P_1R_1}$, is half of what we need to understand Robinson's description of the transition points.

\begin{onlineOnly}
    \begin{center}
\desmos{7rvj2du0mv}{900}{600}   %yfoczp8y2g
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/7rvj2du0mv}{Hidden Trammel Explained}


%Trammel comes from \emph{trēmaculum}, a concatenation of the Latin \emph{trēs} (``three") and \emph{macula} (``mesh"). In middle English it referred to a three-layered fishing net and it has evolved to mean any restraint or hindrance (CITE A SOURCE HERE). And restraining two points on a thin rod to slide along two (usually perpendicular) lines is exactly what we need to force any third point on that rod (or any point fixed in the reference frame of the rod) to trace an ellipse. %that we can draw an ellipse by tracing a third point $P^*$ on the rod. %(more generally, as the endpoints of a segment slide along two perpendicular lines, a point fixed in the reference frame of the segment sweeps out an ellipse). 

%A trammel of Archimedes hides in Figure ??. As the smaller circle ${\cal C}_1$ rolls around the inside of ${\cal C}_2$, the endpoints of the diameter $\overline{PR}$ through the tracing point slide along the axes of the ellipse. This trammel, with arm $\overline{PR}$, is half of what we need to understand Robinson's description of the transition points.

%\begin{onlineOnly}
 %   \begin{center}
%\desmos{nhapbvzxnu}{900}{600}   %yfoczp8y2g
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/nhapbvzxnu}{Wrapping Cycloids Around Circles Ellipse 2BBB}

The other half is another trammel (Figure ??). It slides along the same axes as the the one in Figure ?? and traces the same ellipse with the same tracing point. But now the tracing point lies \emph{between} the pins, at distances $P^* R_2 = P^* R_1 = b+a$ and $P^* P_2 = P^* P_1 = b-a$ from them. So the two arms cut the axes at the same angle ($\angle R_1 \cong\angle R_2$ and $\angle P_1\cong \angle P_2$) and rotate at the same rate as $P^*$ traces the ellipse. But while the arm of the first trammel is never tangent to the ellipse, the arm of the second is tangent four times. And the four points of tangency are the transition points separating the same and opposite-side arclengths of the ellipse.




\begin{onlineOnly}
    \begin{center}
\desmos{rheveebavy}{900}{600}   %yfoczp8y2g
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/rheveebavy}{Double Generation Ellipse 4}

Figure ?? helps explain why with two new circles. The first, ${\cal C}_2$, lies on diameter $\overline{P_2R_2}$ and has radius $b$. The second, ${\cal C}_2^\prime$, is centered at $O$ and has radius $2b$. Set the figure in motion and you'll see ${\cal C}_2$ rolling around the inside of ${\cal C}_2^\prime$ as ${\cal C}_1$ rolls around ${\cal C}_1^\prime$. The point $P^*$, at rest in the reference frames of both rolling circles, traces the ellipse as a roulette in two ways; as a point attached $b=C_1P^*$ units from the center of ${\cal C}_1$ (with radius $a=C_1Q_1$) and as a point $a=C_2P^*$ units from the center of ${\cal C}_2$ (with radius $b=C_2Q_2$).

\begin{onlineOnly}
    \begin{center}
\desmos{2gnd3kgvyz}{900}{600}   %yfoczp8y2g
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/2gnd3kgvyz}{Double Generation Ellipse 6}

We have some details to fill in, but Robinson's Lemma follows from three observations about the animation:

\begin{enumerate}
\item Radius $\overline{C_1Q_1}$ and trammel arm $\overline{P_2R_2}$ are always parallel.

\item The tracing point $P^*$ and the instantaneous centers of rotation $Q_1$ and $Q_2$ of the rolling circles are always colinear.

\item Trammel arm $\overline{P_2R_2}$  is tangent to the ellipse exactly when $\overline{Q_2P^*}$ is perpendicular to $\overline{P_2R_2}$.
\end{enumerate}

So at the transition points, when $\overline{Q_1P^*}$ is perpendicular to $\overline{C_1Q_1}$, (a) and (b) imply $\overline{Q_2P^*}$ is perpendicular to $\overline{P_2R_2}$. Then (c) implies $\overline{P_2R_2}$ is tangent to the ellipse. 

Our proof of (a) and (b) rests on the fact that the radius of each rolling circle is half the radius of its partner. For then Equation
 (\ref{Eq:RollAngle}) implies the vertical lines (not shown) through centers $C_1$ and $C_2$ bisect the congruent angles $\angle Q_1C_1R_1$ and $\angle Q_2C_2R_2$. It follows that $\overline{C_1Q_1}\parallel \overline{C_2P^*}$ (proving (a)) and $\overline{C_1P^*}\parallel \overline{C_2Q_2}$. So $\angle P^*C_1Q_1\cong \angle Q_2C_2P^*$ and $\Delta P^* C_1Q_1 \cong \Delta Q_2C_2P^*$ (SAS). Thus $\angle C_1P^*Q_1 \cong \angle C_2Q_2P^*$ and $\overline{Q_2P^*}\parallel \overline{P^*Q_1}$ (proving (b)).




To prove (c), we focus our attention on the second trammel. With the coordinate system of Figure ?? and $\phi$ the marked angle between the arm and the $x$-axis, $P^*$ has coordinates 
\[
     (x,y) = ((b+a)\cos\phi, (b-a)\sin\phi) , 0\leq \phi <2\pi. 
\]

\begin{onlineOnly}
    \begin{center}
\desmos{6cjnmo1typ}{900}{600}   %yfoczp8y2g
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/6cjnmo1typ}{Double Trammel}

This shows $P^*$ traces an ellipse with semi-major/minor axes $P^*R_2=b+a$ and $P^*P_2 = b-a$. From here we can take either an algebraic or a geometric approach. In either case, the key observation is the same - the sliding arm is tangent to the ellipse exactly when $P^*$ moves parallel to the arm. I like the algebraic approach for a calculus class (where this problem is a variation of the one that relates the speeds of the ends of a sliding ladder), but here we'll take a geometric approach.% by appealing to the instantaneous center of rotation.


The key idea is to find the arm's instantaneous center of rotation. This is the point in the reference frame of the sliding arm momentarily at rest relative to the $xy$-plane in Figure ??. To find this center physically, place two adjacent vertices $P_2$, $R_2$ of a rectangular paper on adjacent edges of a tabe top (Figure ??) (think of segment $\overline{P_2R_2}$ as the trammel's arm and the adjacent edges as the trammel's axes). Now rotate the paper through a small angle about a random point $W$ and points $P_2$ and $R_2$ almost certainly move off the table. But choose the correct point $Q_2$ (the arm's instantaneneous center of rotation) and the trammel's endpoints move along circular paths tangent to the table's edges. This center of rotation is the point where the lines through $P_2$ and $R_2$ normal to the edges (and therefore normal to the velocities of $P_2$ and $R_2$) intersect.

\begin{onlineOnly}
    \begin{center}
\desmos{hi6ug11jbn}{900}{600}   %yfoczp8y2g
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/hi6ug11jbn}{Center of Rotation 3}

Figure ?? shows the velocities of several points on the sliding trammel as it rotates counterclockwise. Because these velocities are perpendicular to the rotation arms from $Q_2$, the foot of the perpendicular from the center of rotation $Q_2$ to $\overline{P_2R_2}$ is the only point moving parallel to the trammel. And because the tracing point $P^*$ lies between $P_2$ and $Q_2$, this proves (c) and completes our proof of Robinson's lemma.



\begin{onlineOnly}
    \begin{center}
\desmos{nnmsmjmmnu}{900}{600}   %yfoczp8y2g
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/nnmsmjmmnu}{Double Trammel 3}


{\bf Remarks:}

\begin{enumerate}
\item Plotting the centers of rotation in the reference frames of the trammel and the $xy$-coordinate systems gives two curves, respectively called the \emph{moving} and \emph{fixed centrodes}. For the sliding trammel, both centrodes are circles. Because $OQ_2 = P_2R_2 = 2b$, the fixed centrode is a circle of radius $2b$ centered at $O$. And because $m\angle R_2Q_2P_2  =\pi/2$, the rolling centrode is a circle on diameter $\overline{P_2R_2}$. A general theorem says that for any smooth motion of a planar body (like our sliding trammel), the rolling centrode rolls without slippoing on the fixed centrode as it carries the planar body along with it (see Hilbert). 

\item One of my favorite problems is the discrete version of constructing the instantaneous center of rotation. In Figure ??, this amounts to finding the one point that represents the same location in the maps below. See Hilbert again, or \emph{Excursions in Geometry??}.

\begin{onlineOnly}
    \begin{center}
\desmos{cthwmjxkp2}{900}{600}   
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/cthwmjxkp2}{Washington Maps}


\item The transition points of the ellipse have an equivalent description as the points of the ellipse that minimize the length of the tangent segments.

%\item Robinson's lemma also applies to the $(a,b,2a)$-epicycloid of Figure ??, where $P^*B = b+a$ and $P^*A=b-a$. This is because the epicyloid and the $(a,b,-2a)$ ellipse share the same transition points and tangent segments at these points.
 
%\begin{onlineOnly}
%    \begin{center}
%\desmos{mdmsv1cqu3}{900}{600}   
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/mdmsv1cqu3}{Ellipse and Nephroid 33}

\item Tracing an ellipse as a hypocyloid in two ways is a special case of the Bernoulli-Euler double generation theorem. This theorem describes how to trace any hypo or epicyloid a second way. Reference needed here.
\end{enumerate}




%\begin{onlineOnly}
%    \begin{center}
%\desmos{q6fiohnpmo}{900}{600}   %yfoczp8y2g
%\end{center}
%\end{onlineOnly}

\href{https://www.desmos.com/calculator/q6fiohnpmo}{Double Generatation Ellipse 5}



 %(The Wilderness Act of 1964 defined wilderness as ``an area where the earth and its community of life are untrammeled by man, where man himself is a visitor who does not remain")


%\begin{onlineOnly}
%    \begin{center}
%\desmos{nfmfnkp9nb}{900}{600}   %yfoczp8y2g
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/nfmfnkp9nb}{Double Generation Ellipse}


%Figure ?? (in the previous section) suggests how to generate an ellipse as a \emph{glissette} (from the French, \emph{glisser}, to slide). It is the curve traced by a point on a segment with a fixed length $L$ as its endpoints slide along two perpendicular lines. In Figure ?? the tracing point lies beyond the endpoints of segment $\overline{AB}$, but still on the line $AB$. Having the tracing point is outside the segment like this makes it possible to build an ellisograph (the trammel of Archimedes), as Tom Silva does in the episode of \emph{This Old House}.





%\section{Conclusion}

%\begin{onlineOnly}
%    \begin{center}
%\desmos{j1fxecamkh}{900}{600}   
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/j1fxecamkh}{Ellipse and Nephroid}


%\section{Double Generation Theorem}
%Double Generation:

%\begin{onlineOnly}
%    \begin{center}
%\desmos{fidce8bipo}{900}{600}   
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/fidce8bipo}{Ellipse as Glissette}



\section{Curtate Cycloids}

The challenge in seeking an arclength theorem for curtate cycloids (see Figure in the introduction) was to find a substitute for the inclination angle. For when the tracing point lies inside a circle rolling on a line, the tangents to the roulette oscillate between two extremes instead of making complete revolutions. This makes it impossible to use the inclination angle to pair points of a curtate cycloid with \emph{all} the points of an ordinary cycloid.


%Had it occured to me to pair points of the curtate cycloid at the same inclination angle, I might have discovered the global theorem illustrated in Figure ??. But even so, a local version of the theorem would probably not have been apparent.

%It did not even occur to me to pair points of the curtate cycloid at the same inclination angle. This pairing might have suggested the global theorem illustrated in Figure 2; that the difference in the concave up and concave down arclengths of the curtate cycloid would equal the arclength of the sister cycloid. But even with this, a local version of such a theorem might have remained elusive. 

%Instead, the double-generation animation of Figure ?? pointed a way forward. Most of all, since there are no curtate or prolate ellipses, the animation suggests the apparent the differences between curtate and prolate cycloids are really superficial. More specifically, tracing an ellipse in two ways as a hypocycloid implies both a global and local theorem equating the arclengths of the $(a,b)$ prolate cycloid (traced by a point $b>a$ units from the center of a circle with radius $a$ rolling on a line) and the $(b,a)$ curtate cycloid. 

%Instead it was the double-generation animation of Figure ?? that suggested an arclength theorem for curtate cycloids.

The double-generation animation of Figure ?? suggests a way forward. Most of all, since there are no curtate or prolate ellipses, the animation suggests the apparent the differences between curtate and prolate cycloids are superficial. More specifically, tracing an ellipse in two ways as a hypocycloid implies both a global and local theorem equating the arclengths of the $(a,b)$ prolate cycloid (traced by a point $b>a$ units from the center of a circle with radius $a$ rolling on a line) and the $(b,a)$ curtate cycloid. 

The key is the scaling factor. Remember for a hypocylcloid, arclengths get halved in switching the fixed centrode from a line to a circle with twice the radius of the rolling circle. So since arclengths between the same rotation angles of the two identical hypocycloids in Figure ?? are equal, arclengths between the same rotation angles of the $(a,b)$-prolate cycloid and the $(b,a)$-curtate cycloid are also equal. 

%But we can still pair points of a curtate cycloid with the same inclination angle. This correspondence pairs each point on the concave-up side ($0\leq \theta < \theta_0 = \arccos(-a/b)$) of a curtate cycloid with a point on the concave-down side ($\theta_0 < \theta \leq \pi$) and the inflection point (where $\theta = \theta_0$ and $\alpha = \arcsin(a/b)$ is a maximum) with itself. On the static circle, the correspondence pairs the endpoints $P_4$ and $P_4^\prime$ of the chords through $Q_4$. So instead of refering to the concave-up and concave-down sides of the cylcoid (which depend on our viewpoint), we refer instead to the far and near sides, traced respectively by the points $P_4$ and $P_4^\prime$ farther from and nearer to $Q_4$. This same description applies also to the prolate cycloid, where the points $P_1$ and $P_1^\prime$ in Figure ?? are respectively farther and nearer to $Q_1$.


%The double-generation theorem of Figure ?? suggested a way forward. Tracing one ellipse as a hypocycloid in two ways implies both a global and local theorem equating arclengths of the $(a,b)$ prolate cycloid and the $(b,a)$ curtate cycloid. The key is the scaling factor. Remember, for a hypocylcloid arclengths get halved in switching the fixed centrode from a line to a circle with twice the radius of the rolling circle. And since arclengths between the same rotation angles of the two identical hypocycloids in Figure ?? are equal, arclengths between the same rotation angles of an $(a,b)$-prolate cycloid and the $(b,a)$-curtate cycloid are also equal. 

In hindsight, we did not need the double-generation theorem to see this. The circles in the animated version of Figure ?? share the same rotation angle $\theta$ and rotate at the same rate. The upper circle has radius $a$ and tracing distance $b$, the lower radius $b$ and tracing distance $a$. Since 
\[
  \angle P_1C_1Q_1 = \angle P_4C_4Q_4 = \pi - \theta,  
\]
$\Delta P_1 C_1 Q_1\cong \Delta Q_4 C_4 P_4$. So the rotation arms $Q_1P_1$ and $Q_4P_4$ are equal and $P_1$ and $P_4$ traverse equal differential arclengths as the circles turn through the differential angle $d\theta$. Thus, the arclengths $s_1= A_1P_1$ and $s_4=A_4P_4$ are equal.

%The two circles, ${\cal C}_1$ with radius $a$ and tracing distance $b$, and ${\cal C}_4$ with radius $b$ and tracing distance $a$, have equal rotation angles (the marked angles at centers $C_1$, $C_2$). Since $\Delta C_1 Q_1 P\cong \Delta C_4P_4Q_4$, the cycloids have equal $Q_1P_1 = Q_4P_4$ rotation arms. Hence, the tracing points $P_1$ and $P_4$ traverse equal differential arclengths as the circles turn through the angle $d\theta$. 

\begin{onlineOnly}
    \begin{center}
\desmos{s6agxwwc4s}{900}{600}   
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/s6agxwwc4s}{Curtate Cycloid ArcLength 35}


%An arclength theorem for curtate cycloids now follows as a corollary to our arclength theorem for prolate cycloids. We just need to check at pairs of rotation angles $0\leq \theta < \theta^\prime \leq \pi$ where the inclination anlges of the curtatate cycloid are equal, the inclination angles of the prolate cycloid are supplementary. This is the content of the following lemma.

%We just need the following lemma.% follows almost directly from this equality of arclengths. now  rests on the following lemma.

We can use this equality of arclengths to translate the arclength theorem for prolate cycloids into one about curtate cycloids. The key is to relate the rotation and inclination angles of the two curves. 

\begin{theorem22}
For lengths $b>a$, let $\phi$ and $\alpha$ respectively be the inclination angles of the $(a,b)$-prolate cycloid and the $(b,a)$-curtate cycloid at the rotation angle $\theta$, $0\leq \theta \leq \arccos(-a/b)$. Then the cycloids have respective inclination angles $\pi - \phi$ and $\alpha$ at rotation angle $\theta^\prime = \pi - \phi + \alpha$. 
\end{theorem22} 


\begin{proof}
%Let $P_4, P_4^\prime$ be points on the curtate cycloid with the same inclination angle $\alpha$ and respective rotation angles $\theta$, $\theta^\prime$, where $0\leq \theta \leq \theta^\prime \leq \pi$. Then in the reference frame of the static circle (radius $b$) where the center of rotation $Q_4$ is at rest, the points $P_4$, $P_4^\prime$, $Q_4$ are colinear and $Q_4P_4 \geq Q_4P_4^\prime$ (Figure ?? (a)). Since $\theta = \angle B_4C_4P_4$ is an exterior angle of $\Delta Q_4 C_4P_4$ , 

Let $P_4$, $P_4^\prime$ be points on the curtate cycloid with inclination angle $\alpha$ and respective rotation angles $\theta$, $\theta^\prime$, where $0\leq \theta \leq \theta^\prime \leq \pi$. Let $P_1$ , $P_1^\prime$ be the points on the prolate cycloid with respective rotation angles $\theta$, $\theta^\prime$ (Figure ?).

\begin{onlineOnly}
    \begin{center}
\desmos{dqd3jjjsfs}{900}{600}   
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/dqd3jjjsfs}{Curtate Cycloid ArcLength 38}

Figure ?? shows these points on the static circles (ie. in the reference frame where the centers of rotation $Q_4$ and $Q_1$ are at rest). 

\[
  \theta  = \phi + \alpha
\] 
and $P_4^\prime$ has rotation angle
\[
   \theta^\prime = \angle B_4 C_4 P_4^\prime =  \theta + (\pi - 2\phi) = \pi - \phi + \alpha  .
\]







\begin{onlineOnly}
    \begin{center}
\desmos{uttndrcxjz}{900}{600}   
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/uttndrcxjz}{Pairings Curtate Cycloid 5}

At the inflection point $T_4$, the points $P_4$ and $P_4^\prime$ coincide (Figure ?? (b) and $\overline{Q_4T_4}$ is tangent to the circle (radius $a$) centered at $C_4$. The inclination angle $\alpha_0 = \arcsin(a/b)$ is a maximum, and the rotation angle has measure $\theta_0 = \arccos(-a/b)$.

Now let $P_1$, $P_1^\prime$ be the points on the prolate cycloid with respective rotation angles $\theta$, $\theta^\prime$. Then in the reference frame of the static circle (radius $a$) where the center of rotation $Q_1$ is at rest (Figure ??), the points $C_1$, $P_4$, and $P_1$ are colinear, as are the points $C_1$, $P_4^\prime$, $P_1^\prime$. Thus, $P_1$ prime has incination angle 
\[
    \angle C_1Q_1P_1^\prime = \theta^\prime - \alpha = \pi - \phi.
\]

\begin{onlineOnly}
    \begin{center}
\desmos{grewshzmd7}{900}{600}   
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/grewshzmd7}{Pairings Curtate and Prolate 5}

\end{proof}

{\bf Remark:} Instead of refering to the concave-up and concave-down sides of a curtate cylcoid (which depend on our viewpoint), we refer instead to the far and near sides, traced respectively by the points $P_4$ and $P_4^\prime$ farther from and nearer to $Q_4$. This same description applies also to the prolate cycloid, where the points $P_1$ and $P_1^\prime$ in Figure ?? are respectively farther and nearer to $Q_1$.


%And the correspondence pairs each point on the concave-up side at rotation angle $0\leq \theta < \theta_0$ with the point on the concave-down side at rotation angle
%\[
  % \theta^\prime = \theta + (\pi - 2\phi) = \pi - \phi + \alpha  .
%\]



%These tracing points $P_4$ and $P_5$ of the curtate cycloid at the same inclination angle $\alpha$. The correspondence pairs the endpoints of chords $\overline{P_4P_5}$ through $Q_4$ (extended). It pairs the inflection point, where 

% It pairs points on the concave-up side ($0\leq \theta < \theta_0 = \arccos(-a/b)$) of the curtate cycloid with points on the concave-down side ($\theta_0 < \theta \leq \pi$) and the inflection point (where 

%Then since a rotation about $C_4$ through the angle
%\[
%    \angle P_4 C_4 P_4^\prime = \pi - 2\phi
%\]
%takes $P_4$ to $P_5$, the curtate cycloid has inclination angle $\alpha$ at rotation angle
%\[
%   \theta^\prime = \theta + (\pi - 2\phi) = \pi - \phi + \alpha .
%\]
%See the upper frame of Figure ??.
%\end{proof}



%\begin{proof}
%Since $\theta$ is an exterior angle of congruent triangles $\Delta P_1 C_1 Q_1$ and $\Delta Q_4 C_4P_4$ (Figure ??), 
%\[
%  \theta  = \phi + \alpha.
%\] 

%Now let $P_4^\prime$ be the second point where the rotation arm $\overline{Q_4P_4}$ intersects the circle of radius $a$ centered at $C_4$. Then a rotation about $C_4$ through the angle
%\[
%    \angle P_4 C_4 P_4^\prime = \pi - 2\phi
%\]
%takes $P_4$ to $P_4^\prime$. So at rotation angle
%\[
%   \theta^\prime = \theta + (\pi - 2\phi) = \pi - \phi + \alpha,
%\]
%the curtate cycloid has inclination angle $\alpha$ as illustrated in the upper frame of Figure ??. And at rotation angle $\theta^\prime$, the prolate cycloid has inclination angle
%\[
 %\angle C_1Q_1P_1 = \theta^\prime - \alpha = \pi - \phi.
%\]
%\end{proof}

%{\bf Remark:} The correspondence $\theta \leftrightarrow \theta^\prime$ pairs each point on the concave-up side ($0\leq \theta < \theta_0 = \arccos(-a/b)$) of a curtate cycloid with a point on the concave-down side ($\theta_0 < \theta \leq \pi$) and the inflection point (where $\theta = \theta_0$ and $\alpha = \arcsin(a/b)$ is a maximum) with itself. On the static circle, the correspondence pairs the endpoints $P_4$ and $P_4^\prime$ of the chords through $Q_4$. So instead of refering to the concave-up and concave-down sides of the cylcoid (which depend on our viewpoint), we refer instead to the far and near sides, traced respectively by the points $P_4$ and $P_4^\prime$ farther from and nearer to $Q_4$. This same description applies also to the prolate cycloid, where the points $P_1$ and $P_1^\prime$ in Figure ?? are respectively farther and nearer to $Q_1$.



%$One the concave-up side, the tracing point and the center of the rolling circle lie on the same side of the tangent to the cycloid (represented by the dotted line through $P_4$ normal to $\overline{P_4Q_4}$ in Figure ??). On the concave-down side these two points lie on opposite sides of the tangent line. And at an inflection point (point $P$ in Figure ??), the center and tracing point lie on the tangent line.

%\begin{onlineOnly}
%    \begin{center}
%\desmos{5tmoep2gja}{900}{600}   
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/5tmoep2gja}{Pairings Curtate and Prolate Cycloids}

%So defining the \emph{same-side}


The law of sines in $\Delta P_4 C_4Q_4$ in Figure ?? and the arclength theorem for prolate cycloids, now imply the following.

\begin{onlineOnly}
    \begin{center}
\desmos{g7cb6r89xl}{900}{600}   
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/g7cb6r89xl}{Curtate Cycloid Inflection Points}


\begin{theorem23}
%Let $s_1 = s_1(\alpha)$ and $s_2 = s_2(\alpha)$ be respectively the same and opposite-side arclengths of an $(a,b,r)$ curtate ($a>b$) epi or hypocylcoid between inclination angles $0$ and $\alpha$ , $0< \alpha \leq \arccos(-b/a)$. Let $s_3 = s_3(\phi)$ be the arclength of the sister $(a,a,r)$ epicycloid (resp. hypocylcoid) curve between inclination angles $0$ and $\arcsin((b/a)\sin\alpha)$. Then
%\[
%   s_1 - s_2 = s_3 .
%\]

\begin{enumerate}
\item For lengths $b>a$,  inclination angles $0\leq \phi \leq \pi/2$ and
\[
    \alpha = \arcsin\left( \frac{a}{b}\sin\phi  \right) ,
\]
and rotation angles $0\leq \theta\leq \arccos(-a/b)$, the following are equal:

\begin{enumerate}
\item the arclength of the ordinary cycloid (radius $a$) between inclinations angles $0$ and $\phi$

\item the difference in the far-side ($0\leq \theta \leq \phi + \alpha$) and near-side ($\pi- \phi + \alpha \leq \theta \leq \pi$) arclengths of the prolate $(a,b)$-cycloid between inclination angles $0$ and $\phi$

\item  the difference in the far-side ($0\leq \theta \leq \phi + \alpha$) and near-side ($\pi- \phi + \alpha \leq \theta \leq \pi$) arclengths of the curtate $(b,a)$-cycloid between inclination angles $0$ and $\alpha$ .

\end{enumerate}

\item The same ????

\end{enumerate}

\end{theorem23}

Figure ?? shows an example, where with arclengths $s_i = A_iP_i$, $i=1,\ldots , 5$, 
\[
     s_1  -  s_2 = s_3 = s_4 -  s_5.  
\]

\begin{onlineOnly}
    \begin{center}
\desmos{opaijymtyr}{900}{600}   
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/opaijymtyr}{Curtate Cycloid ArcLength 5}



%\begin{onlineOnly}
%    \begin{center}
%\desmos{ee9rmqgmbw}{900}{600}   
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/ee9rmqgmbw}{Curtate Cycloid ArcLength 4B}



%\begin{onlineOnly}
%    \begin{center}
%\desmos{d1kzu52vf4}{900}{600}   
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/d1kzu52vf4}{Curtate Cycloid ArcLength 36}



Finally, a similar equalities of arclengths hold for epi and hypocycloids. Figure ?? illustrates the global version.

\begin{onlineOnly}
    \begin{center}
\desmos{xxybrtbt0o}{900}{600}   
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/xxybrtbt0o}{Cardioid with Curtate and Oblate EpiCycloids 2}

{\bf Remark:} Euler ...


%There appears to be an arclength theorem for curtate cycloids that might follow from the double generation theorem.

%Global Statement: When $b<a$ the curtate $(a,b)$-cycloid has inflection points. Then the upper arclength (above the inflection point) exceeds the lower arclength by $8b$, the length of the ordinary $(b,b)$-cycloid. Note this difference is independent of $a$. Compare with Robinson's theorem on prolate cycloids. 

%\begin{onlineOnly}
%    \begin{center}
%\desmos{h5df8repfp}{900}{600}   
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/h5df8repfp}{Curtate Cycloid Arclength 3}


%Here's another version of the worksheet.

%\begin{onlineOnly}
%    \begin{center}%
%\desmos{rytjk2xes8}{900}{600}   
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/rytjk2xes8}{Curtate Cycloid Arclength 4}




%\begin{onlineOnly}
 %   \begin{center}
%\desmos{blx43t49kj}{900}{600}   
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/blx43t49kj}{Ellipse Doubly Generated 2}



\end{document}
