\documentclass{ximera}
\title{Cycloidal arc lengths: An Animated Approach, Abstract}

\newcommand{\pskip}{\vskip 0.1 in}

%\newtheorem{theorem19}{Robinson's arc length Theorem for Prolate Cycloids, Local Version}
%\newtheorem{theorem20}{An arc length Theorem for Prolate Epi and Hypocycloids}
%\newtheorem{theorem21}{Robinson's Lemma}
%\newtheorem{theorem22}{The Inclination Lemma}
%\newtheorem{theorem23}{An arc length Theorem for General Cycloids}

\begin{document}
\begin{abstract}
\end{abstract}
\maketitle
 
\begin{exploration} \label{SlidingLadders}
There is exactly one point $P$ on the quarter-ellipse
\[
 \frac{x^2}{u^2} + \frac{y^2}{v^2} = 1 \, , \, x,y\geq 0,
\]
where the tangent segments $\overline{PA}$ and $\overline{PB}$ from $P$ to the coordinate axes have lengths $PA=v$ and $PB=u$ equal to the semi-axes. This point divides the quarter-ellipse into arc lengths $PX=s_x$ and $PY=s_y$ whose difference $s_y - s_x$ is equal to the difference  $u-v$ in the semi-axes.

%At this point the difference $s_y - s_x$ in the arc lengths $PY$ and $PX$ of the ellipse from $P$ to the coordinate axes equals the difference $u-v$ in the lengths of the semi-axes. This paper is partly about proving this theorem. But it is mostly about taking a local approach to this and other calculus problems.

The first step in understanding why is to animate the figure, either by dragging the slider $\phi$ in Line 2 or pressing the play button. Do so and you'll see the endpoints of segment $\overline{AB}$ and another segment $\overline{CD}$ slide along the coordinate axes as the point $P$, attached to the longer segment $\overline{AB}$ (length $u+v$) and to an extension of the shorter $\overline{CD}$ (length $u-v$), sweeps out the quarter-ellipse. 

You can drag the sliders $a=(u-v)/2$ (Line 7) and $b=(u+v)/2$ (Line 8) to change the shape and size of the ellipse.


\begin{onlineOnly}
    \begin{center}
\desmos{b8xr2fpea3}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/b8xr2fpea3}{Cylcoid Abstract Sliding Ladder}
\end{exploration}


\begin{exploration} \label{LadderwithCircles}
This animation gives another perspective of the motion in the first. Here each sliding segment is a fixed diameter of a circle rolling around the inside of another with twice the radius. The circles each roll once through their circumferences as the point $P$ traces a semi-ellipse. 

The animation begins at the point in the second quadrant where $\overline{AB}$ is tangent to the ellipse and the tracing point $P$ crosses the rolling tangent line ${\cal L}_2$. 

%The circle on diameter $CD=2a = u-v$ helps establish the arc length equality in Theorem 1. The circle on diameter $AB = 2b = u+v$ helps identify the transition point separating the negative and positive parts of the arc length.

Play the slider $\phi_2$ in Line 2 to start the motion.
%In this animation the sliding segments $\overline{AB}$ and $\overline{CD}$ are each fixed diameters of circles rolling around the inside of other circles another with twice the radius. Each circle rolls once through its circumference as the point $P$ traces a semi-ellipse. The circle on diameter $CD=2a = u-v$ helps establish the arc length equality in Theorem 1. The circle on diameter $AB = 2b = u+v$ identifies the transition point geometrically. %Each rolls once through its circumference as they rotate at the same constant rate.

\begin{onlineOnly}
    \begin{center}
\desmos{wcvw7kjsdd}{900}{600}% wcvw7kjsdd  %exj0ey5hjf  b9xvjd6s16
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/wcvw7kjsdd}{Cylcoid Abstract Sliding Ladders with Circles 2}
\end{exploration}


\begin{exploration}  \label{EllipseProlateCycloid}
The second figure below shows the paths traced by the points $P$ and $C$ in the reference frame of the rolling tangent line ${\cal L}_2$ in Exploration 2. One is an arch of a prolate cycloid, traced by the point $P$ outside the rolling circle (radius $a = (u-v)/2$) at a distance $b=(u+v)/2$ from its center. The other is an arch of an ordinary cycloid, traced by the point $C$ on the rolling circle. Play the slider $\phi_2$ in Line 2 to complete the paths and see the circle on diameter $\overline{CD}$ roll in the reference frames of the fixed circle (top) and the rolling tangent line (bottom).

%Because the rolling circle has half the radius of the fixed, it rotates half as fast in the top frame as it does in the bottom. So any point at rest in the reference frame of the rolling circle moves half as fast in the top frame as it does in the bottom. Thus,

%Because the circle rotates twice as fast in the second frame as it does in the first, any point at rest in the reference frame of the rolling circle moves twice as fast in the second frame as in the first. So at any stage of the motion, the length $A_2P_2$ of the prolate cycloid is twice the corresponding length $AP$ of the ellipse and the length $Z_2C_2$ of the ordinary cycloid is twice the length of segment $\overline{ZC}$.

%Because the bottom circle rotates twice as fast as the top, a point at rest in the reference frame of the rolling circle moves twice as fast in the reference frame of the rolling tangent as it does in the reference frame of the fixed circle. So at any stage of the motion, the length $X_2P$ of the prolate cycloid is twice the corresponding length $XP$ of the ellipse and the length $Z_2C$ of the ordinary cycloid is twice the length of segment $\overline{ZC}$.

Because the bottom circle rotates twice as fast as the top, a point fixed in the reference frame of the rolling circle moves twice as fast in the top reference frame as it does in the bottom. So at any stage of the motion, the length $X_2P$ of the prolate cycloid is twice the corresponding length $XP$ of the ellipse and the length $Z_2C$ of the ordinary cycloid is twice the length of segment $\overline{ZC}$.

The difference $2s_1-2s_2$ in the lengths of the prolate cycloid above and below ${\cal L}_2$ equals the length $2s_3$ of the ordinary cycloid (Theorem 2).  

%In particular,
%\[
%  \text{arc}(XP) =0.5 \text{arc}(X_2P_2)  \hskip 0.3 in \text{and} \hskip 0.3 in ZC = \frac{1}{2}\text{arc}(Z_2C_2).
%\]

\begin{onlineOnly}
    \begin{center}
\desmos{jrcqvu50gh}{900}{600}% 1tuo4fouw8
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/jrcqvu50gh}{Ellipse and Prolate Cycloid 3}

\end{exploration}

\iffalse
******************************************************************************************************

\begin{exploration} \label{CurtateProlateCyloidsRolling}
This figure shows the paths traced by the points $P$ in the reference frames of the rolling tangent lines of Exploration 2. The first path is one arch of a curtate cycloid, where $P$ is inside the rolling circle (radius $b = (u+v)/2$), at a distance $a=(u-v)/2$ units from its center. The second, dual to the first, is one arch of the prolate cycloid of Exploration 3, where $P$ is outside the rolling circle (radius $a$), $b$ units from its center. The third is an ordinary $a$-cycloid, traced by a point on a rolling circle with radius $a$. 

Proving Theorem 1 amounts to showing the arc lengths $\sigma_1, \sigma_2$ of the prolate cycloid above and below ${\cal L}_2$ respectively equal the arc lengths $s_1$, $s_2$ of the concave-up and concave-down arcs of the curtate cycloid, and that the difference $\sigma_1 - \sigma_2 = s_1-s_2$ equals the arc length $s_3$ of the $a$-cycloid.

Play the slider $u$ in Line 1 to see the circles roll in the reference frames of the rolling tangent lines ${\cal L}_1$ and ${\cal L}_2$. The first two circles rotate at the same rate and because the arc lengths $A_1P_1$ and $A_2P_2$ are each twice the arc length $XP$ of the semi-ellipse in Exploration 2,  $A_1P_1=A_2P_2$. 

\begin{onlineOnly}
    \begin{center}
\desmos{b9xvjd6s16}{900}{600}  % fw0cfv3wk3 b9xvjd6s16
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/b9xvjd6s16}{Cylcoid Abstract Prolate Curtate Cycloids Two Frames}
\end{exploration}

***************************************************************************************************
\fi


\begin{exploration} \label{CurtateProlateCycloids}

This animation includes the path traced by $P$  in the reference frame of the rolling tangent line ${\cal L}_1$ in Exploration 2. It is one arch of a curtate cycloid, traced by the point $P$ inside the rolling circle (radius $b = (u+v)/2$) at a distance $a=(u-v)/2$ from its center. Since $P$ rotates at the same rate in the reference frames of the two tangent lines, the arclengths $X_2P$ and $X_3P$ are always equal. Play the slider $\phi_2$ in Line 2 to start the motions.

\begin{onlineOnly}
    \begin{center}
\desmos{aotcfo08mn}{900}{600}    %lbq2agtef6
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/aotcfo08mn}{Ellipse and Prolbat Cylcoid Abstract 2}
\end{exploration}



\begin{exploration} \label{Theorem2Local}
This figure illustrates the local version of Theorem 2 (Exploration 3). Here $s_1 = A_2P$ is the length of the prolate cycloid above ${\cal L}_2$ between inclination angles $0$ and $\phi$, $s_2= A_2^\prime P^\prime$ the length of the same curve below ${\cal L}$ between the same inclination angles, and $s_3 = A_3C$ the length of the ordinary cycloid between these inclination angles. 

The equality $s_1 - s_2 = s_3$ of arc lengths holds for any inclination angle $\phi$, $0\leq \phi \leq \pi/2$. Drag the slider $\phi$ in Line 1 to change the inclination angle.



% and $s_2=A_2^\prime P^\prime$ are the lengths of the prolate cycloid above and below ${\cal L}_2$ between inclination angles $0$ and $\phi$. And $s_3=A_3P_3$ is the length of the ordinary cycloid between these same inclination angles. 

\begin{onlineOnly}
    \begin{center}
\desmos{i6scaz6pss}{900}{600}    %lbq2agtef6
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/i6scaz6pss}{Cylcoid Abstract Prolate Curtate Local 5}
\end{exploration}


\begin{exploration} \label{CenterofRotation}

Imagine two coincident planes moving relative to one another. If the relative motion is not a translation, each plane has a unique point momentarily at rest relative to the other plane.   %, like the reference frames of the rolling circle and the line ${\cal L}_2$ in Exploration 3. If the relative motion is not a translation, there each plane has a unique point momentarily at rest relative to the other plane. 

For the reference frame of the rolling circle in Exploration 3 for example, this \emph{instantaneous center of rotation} is the point of the circle in contact with the line ${\cal L}_2$. A mathematician's proof might begin by parameterizing the cycloid traced by a point attached to the rolling circle. But for an engineer or a physicist the explanation is simply the definition of what it means for a wheel to roll without slipping; there is no relative motion at the point of contact.

A differential displacement of each plane relative to the other is a rotation about the center of rotation. In the example below, the planes are the table top (stationary from our perspective) and the rectangular paper as its adjacent corners $A$ and $B$ slide along the table's  bottom and left edges.  Drag the slider $\beta$ in Line 1 of the worksheet to rotate the paper through a small angle about the point $W$ and you'll see the corners move off their edges. But when $W$ is the paper's instantaneous center of rotation, a small rotation about $W$ moves these corners along circular paths tangent to the edges. Try dragging $W$ to find this center.

\begin{onlineOnly}
    \begin{center}
\desmos{demmbopxqd}{900}{600}    %lbq2agtef6
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/demmbopxqd}{Center of Rotation Abstract}
\end{exploration}




\begin{exploration} \label{Erp3r}
The centers of rotation trace curves in each plane and rolling one of these curve on the other reproduces the relative motion of the planes. Play the slider $\phi$ in Line 1 of the animation below to see these curves for the sliding paper.


\begin{onlineOnly}
    \begin{center}
\desmos{ng2ltiwmnx}{900}{600}    %lbq2agtef6
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/ng2ltiwmnx}{Center of Rotation Abstract 2}
\end{exploration}

\end{document}
