\documentclass{ximera}
\title{Calculus and the Quadratic Formula}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
We use calculus to gain insight into the quadratic formula.
\end{abstract}
\maketitle

 We offer several interpretations of the quadratic formula from the perspectives of differential and integral calculus.

\section*{Subnormals and the Quadratic Formula}
Our first interpretation came about in checking that the quadratic formula is dimensionally correct and reveals a property of parabolas hidden in the formula itself.

%\begin{exercise}  
%  One of the following expressions gives the area bounded by the $x$-axis and the parabola $y=ax^2+bx+c$. Which one?  
%  \begin{multipleChoice}  
%    \choice{$(b^2-4ac)^{3/2}$}  
%    \choice[correct]{$\frac{(b^2-4ac)^{3/2}}{12a^2}$}  
%    \choice{$a^2(b^2-4ac)^{3/2}$} 
%    \choice{$\frac{(b^2-4ac)^{3/2}}{a^3}$} 
%   \choice{None of the above}
%\choice{All of the above}
%  \end{multipleChoice}  
%\end{exercise}  

%\begin{exploration}\label{exp:orth2}
%Let $W =\mbox{span}\left({\bf v}_1,{\bf v}_2\right)$. $W$ is a plane through the origin in $\mathbb{R}^3$.  Use the navigation bar at the bottom of the interactive window to go through the steps of constructing an orthogonal basis for $W$.  RIGHT-CLICK and DRAG to rotate the image for a better view.
 
%\pdfOnly{
%Access GeoGebra interactives through the online version of this text at
 
%\href{https://ximera.osu.edu/oerlinalg}{https://ximera.osu.edu/oerlinalg}.
%}
 
%\begin{onlineOnly}
%    \begin{center}
%\geogebra{zghsfkym}{900}{600}
%\end{center}
%\end{onlineOnly}
%\end{exploration}



%Units should play a more prominent role in our classes, particularly in places where they seem not to play any role at all. The quadratic formula, for example, comes without units. But in any application the coefficients have meaning and it pays to at least verify the formula is dimensionally correct. And this kind of dimensional thinking can lead to some unexpected insights. %might change the way you look at the quadratic formula.

In its most basic application, the quadratic formula
\[
   x = -\frac{b}{2a}\pm \frac{\sqrt{b^2-4ac}}{2a}
\]
describes how to locate the $x$-intercepts of the parabola 
\begin{equation}
 y=ax^2+bx+c.   \label{Eq:Parabola}
 \end{equation}
 The coordinates $x,y$ and the constant term $c$ have dimensions of length. The coefficient $b$ is dimensionless and gives the slope of the tangent line at the $y$-intercept $Q(0,c)$. Since the product $ax^2$ is a length, so is the reciprocal $a^{-1}$ of the leading coefficient. This reciprocal has several interpretations, each describing the parabola's size. One is related to the parabola's radius of curvature $(2a)^{-1}$ at the vertex and another to its focal length, $(4a)^{-1}$. 

But a third interpretation, our focus here, is encoded in the formula and plays a key role in the solution to the following problem.

\begin{exercise}
Let $P$ be a point on the parabola $y=ax^2$ different from the origin $O$. Let $Q$ be the point where the normal lines to the parabola at $P$ and $O$ intersect. Find the limiting position of $Q$ as $P\to O$.

\begin{exploration}
First guess what happens to $Q$ as $P$ approaches $O$. Then drag the slider $x_0$ in Line 2 to zero to see if you were correct.
\begin{onlineOnly}
    \begin{center}
\desmos{wuazgkeukm}{450}{600} 
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/wuazgkeukm}{151: Normals to Parabola 2}
\end{exploration}

\begin{explanation}
I like this problem because the result is surprising. It seems natural to guess $Q$ goes off to infinity because the normal lines at $P$ and $O$ tend toward being parallel. This would be correct for any curve $y=a|x|^p$ with $p>2$, but not for the parabola $y=ax^2$.

For the parabola, because the normal line at $P(x_0,ax_0^2)$ has slope
\[
        m_\perp = -\frac{1}{2ax_0} ,
\] 
the change in the $y$-coordinate in moving from $P$ to $Q$ is %the \emph{constant}
\[
 \Delta y =  m_\perp (\Delta x) =  \left( -\frac{1}{2ax_0}\right)(-x_0)  = \frac{1}{2a} .
\]

So $Q$ has $y$-coordinate
\[
     y_Q = ax_0^2 + \frac{1}{2a}
\]
and as $P$ approaches the origin, $Q$ approaches the point with coordinates $(0,1/(2a))$.
\end{explanation}
\end{exercise}

The surprising part about the above solution is that in moving along the normal line from a point on any parabola to the axis of symmetry, the $y$-coordinate changes by the \emph{constant}
\[
   \Delta y = \frac{1}{2a} .
\]
This fact, what I'll call the \emph{constant subnormal property of parabolas} (see note below) is actually encoded twice in the quadratic formula. 

Just the first part 
\[
    x = -\frac{b}{2a}
\]
of the formula, an equation for the axis of symmetry, proves this property. The key is to realize that in moving from $Q(0,c)$ to the axis of symmetry along the normal line with slope $m_\perp = -1/b$ at $Q$, the $x$-coordinate changes by $\Delta x = -b/2a$ and the $y$-coordinate by
\begin{equation}
 \Delta y = m_\perp \Delta x = \left( -\frac{1}{b}\right) \left(- \frac{b}{2a}  \right)= \frac{1}{2a} .  \label{Eq:Subnormal}
\end{equation}
But because $\Delta y$ is independent of $b$, (\ref{Eq:Subnormal}) also records the change in the $y$-coordinate in moving from \emph{any} point $P$ of the parabola (\ref{Eq:Parabola}) to the axis of symmetry along the normal at $P$.

%We can prove this same property of parabolas more directly by considering just the parabola $y=ax^2$. The normal at the point $P(x_0, ax_0^2)$ has slope $m_\perp = -1/(2ax_0)$. So in moving from $P$ to the axis of symmetry along the normal, the $y$-coordinate changes by
%\[
%        \Delta y = m_\perp \Delta x = \left( -\frac{1}{2ax_0}\right) \left(-x_0\right)= \frac{1}{2a}   .
%\]

For the same reason, the constant subnormal property is encoded again in the second term of the quadratic formula because the tangents to the parabola at its $x$-intercepts have slopes $\pm \sqrt{b^2-4ac}$. So the $x$-intercepts of the parabola (\ref{Eq:Parabola}), if distinct, are the points where the non-vertical normal lines from the point $(-b/(2a), 1/(2a))$ intersect the parabola.

\emph{Note:} In Leibniz's terminology, segment $PN$ is the \emph{normal} and segment $ON$ the \emph{subnormal} at a point $P$ of the curve $y=f(x)$ in the figure below ($PN$ and $PO$ are respectively normal to the curve and the $x$-axis.) 

\begin{onlineOnly}
    \begin{center}
\desmos{motd0b9nmq}{450}{600} 
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/motd0b9nmq}{151: Leibniz Subnormal}



parabola $ax=y^2$ has a constant \emph{subnormal} $2a^*$. At a point $P$ on a given curve, the subnormal is the \emph{signed} distance $ON$ from $O$, the foot of the perpendicular from $P$ to the $x$-axis, to $x$-intercept of the normal at $P$. %the point $N$ where the normal at $P$ intersects the $x$-axis.

\begin{exploration}\label{exp:cq1}
Drag the slider $b$ to move the parabola $y=ax^2 +bx+c$ and check that the length of segment $AB$ remains unchanged. Also drag the sliders $a$ and $c$ and watch what changes and what remains the same.
 
\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/2apdkp8rwe}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{2apdkp8rwe}{450}{600} \desmos{2apdkp8rwe}{450}{600}
\end{center}
\end{onlineOnly}
\end{exploration}


%In Leibniz's terminology, the parabola $a^* x=y^2$ has a constant \emph{subnormal} $2a^*$. At a point $P$ on a given curve, the subnormal is the \emph{signed} distance $ON$ from $O$, the foot of the perpendicular from $P$ to the $x$-axis, to $x$-intercept of the normal at $P$. %the point $N$ where the normal at $P$ intersects the $x$-axis.
%(Reference: David Dennis, Jere Confrey, Functions of a Curve :
%Leibniz's Original Notion of Functions and its Meaning for the Parabola, \emph{CMJ}, March 1995, v.26 (2), pp.124-130.) 

%This property of parabolas, that they have constant subnormals, is encoded twice in the quadratic formula. Once in the first term as we have just seen, and again in the second term because the slopes of the tangents to the parabola (\ref{Eq:Parabola}) at its $x$-intercepts are $\pm \sqrt{b^2-4ac}$. 

This constant change (\ref{Eq:Subnormal}) in the $y$-coordinate is encoded again in the second term of the quadratic formula because the slopes of the tangents to the parabola at its $x$-intercepts are $\pm \sqrt{b^2-4ac}$. So we can regard the formula as telling us how to locate the $x$-intercepts:

\pskip

a. Start at the $y$-intercept $Q(0,c)$.

\pskip

b. Move along the normal at $Q$ until the $y$-coordinate changes by $1/(2a)$, thereby landing on the axis of symmetry. 

\pskip

c. Move along the axis of symmetry to the point with $y$-coordinate $y=1/(2a)$.

\pskip

d. Move to the $x$-axis, along the lines with slopes $\pm \sqrt{b^2-4ac}$, thereby landing at the intercepts.


\section{Math Department}

Is the quadratic formula dimensionally correct?


For example, suppose the coordinates $x$, $y$ are measured in meters for the parabola 
\[
     y  = ax^2 + bx + c \, , \, c\neq 0.
\]

\begin{enumerate}
\item What are the units of $c$, $b$, $a$?

\item Give geometric interpretations of $c$, $b$, $a$.

\item What are the units of $-b/2a$? Interpet its meaning.

\item What are the units of 
\[
     \frac{\sqrt{b^2-4ac}}{2a}?
\]
Interpet its meaning.

\end{enumerate}

\begin{exploration}\label{exp:cq1}
Drag the slider $b$ to move the parabola $y=ax^2 +bx+c$ and check that the length of segment $AB$ remains unchanged. Also drag the sliders $a$ and $c$ and watch what changes and what remains the same.
 
\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/2apdkp8rwe}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{1kxcom3p2t}{450}{600} \desmos{1kxcom3p2t}{450}{600}
\end{center}
\end{onlineOnly}
\end{exploration}


\href{https://www.desmos.com/calculator/1kxcom3p2t}{Calculus and the Quadratic Formula 22}





\section{Cylindder}


\begin{exploration}\label{exp:orth1}
Cylinder
\begin{center}
\desmos{4rzcqbw4at}{800}{600}
%\desmos{64a5f6cf8761ebbaaa07aea9}{800}{600}
\end{center}
\end{exploration}

%64a5f6cf8761ebbaaa07aea9
%https://teacher.desmos.com/activitybuilder/custom/64359cf956e5d599cc48e5d8

%https://teacher.desmos.com/activitybuilder/custom/64a5f6cf8761ebbaaa07aea9
https://www.desmos.com/calculator/o4zmtwx6hf

\end{document}
